Draft Article

### 2020-01-01 article.tex
Andrew upload

% \section{Spy satellites}

>>>>

\section{Spy satellites}
Gary Powers had been in Soviet prison for less than three months when the United States launched an instrument of espionage far more advanced, and far more damaging, than the U-2 could ever be. Unlike the U-2, the first generation of spy satellite had limitless airspace, no countermeasures, and no technological antecedents; the world had never seen anything like it. In 1955, when the US and the USSR both announced their plans to build an artificial satellite, Outer Space was a true blank slate for international norms.

Only four years after the first U-2 overflight, a familiar dynamic could have played out when the first spy satellite launched in August 1960: the US develops a new photo-reconnaissance technology which its main adversary is unable to prevent; the USSR diplomatically objects to the technology's use; the USSR finally develops a weapon capable of counteracting the new technology; and the USSR employs it. Each one of these steps came to pass except the last---even after developing a feasible anti-satellite (ASAT) weapon, the Soviet Union never made use of it. And the onus was on the Soviets to push back against American efforts to normalize satellite reconnaissance. \emph{Sputnik} briefly put the USSR at the forefront of the final frontier, but the satellite itself was totally innocuous; it collected data about the density of the atmosphere and propagation of radio waves.\footcite{nasa_sputnik_2019} It was the United States that would first launch a satellite capable of photo-reconnaissance---and it was the Soviet Union that allowed it to happen.

Spy satellites are not just one of the best edge cases for norm against shooting down a satellite, they are the reason it exists in the first place.\footnote{I'm not suggesting that, absent spy satellites, nations would have just constantly shot down each other's weather satellites. But remember that the CIA planned a cover for the U-2 in which it was a ``weather plane'' that had gone off course. The USSR forced a cargo plane to land in Hungary because they suspected it of espionage. That never happens in space, because intelligence norms are strong enough that the accusation of dual-use is never even raised. There are no good and bad satellites; as long as a satellite doesn't have literal weapons, then there is no plausible reason to shoot it down.} The reconnaissance satellite succeeded because the Eisenhower administration purposefully designed its space program to establish a norm of peaceful development in space---with the single goal of legitimizing satellites for intelligence-gathering. By protecting the peaceful uses of outer space, the United States finally gained unrestricted access to aerial intelligence of the type that an Open Skies policy might have provided. That all satellites, regardless of purpose, are immune to hostilities as a result is just a happy byproduct.

In this chapter I will explain how the United States developed a norm protecting the use of reconnaissance satellites. Unlike in previous chapters, it is not necessary to prove that there is a muted diplomatic response to satellite espionage because the lack of response is self-evident: no military action has ever been taken against satellites, and the Soviet Union only objected to reconnaissance satellites until it were able to make its own. The first section looks at the 10-year period prior to the first spy satellite, during which the United States built a diplomatic strategy that would legitimize its eventual launch. The section after analyzes the deployment of early spy satellites, and how it altered the security dynamic between both the US and the USSR. The chapter ends by comparing the possible explanations for why the peaceful uses of space norm endured in spite of the threat that it posed to Soviet security.
